Population genetics with evolving mutation processes: inference of mutation spectrum histories from modern genomic variation

### 2019-11-16 manuscript/main.tex
my preferred notation for regularization parameters
acknowledgements
update title and abstract

\title{Coalescent inference of mutation spectrum histories from sample frequency spectra}

\author{
W.S. DeWitt$^{1,4}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1}$\\
{\small
$^1$Department of Genome Sciences,
$^2$Paul G.\ Allen School of Computer Science \& Engineering,}\\
{\small and $^3$Department of Biology, University of Washington, Seattle, WA}\\
{\small $^4$Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
% \small{$^\ast$ Equal contribution}
}

@@

Single nucleotide variant (SNV) frequencies partitioned according to local nucleotide context vary among human ancestry groups and among great ape species, indicating variation and divergence of the mutation process.
The sample frequency spectrum (SFS)---the distribution of derived allele frequencies among sampled haplotypes---is a well-studied population genetic summary statistic that is sensitive to demographic history.
We extend a coalescent framework for demographic inference from the SFS to accommodate inference of mutation spectrum histories from the sample frequency spectrum resolved by $k$-mer nucleotide context ($k$-SFS).
We formulate inference of mutation spectrum histories from the $k$-SFS as a linear inverse problem.

@@

The first option we call ``soft'' rank regularization, with 
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{soft}(Z) 
  = 
  \lambda_\mathrm{spline} 
  \frac{1}{2} \left\|\Delta Z\right\|_2^2
  +
  \lambda_\mathrm{rank} \| Z \|_*.
\end{equation}
The spline term controls the smoothness of solutions, 
and the rank term allows us to recover low-rank solutions.
The rank penalty is the nuclear norm $\| Z \|_*$, 
which is equivalent to the $\ell^1$-norm on the singular values.
If $Z = U S V^\intercal$ is the SVD of $Z$, with $S = \mathrm{diag}(\sigma)$ then $\| Z \|_* = \| \sigma \|_1$.
The parameters $\lambda_\mathrm{spline}$ and $\lambda_\mathrm{rank}$ control the overall strength
of smoothing and rank regularization.
The function $R_\mathrm{soft}$ is convex; therefore we have a convex optimization problem 
in this scenario.

The second option is ``hard'' rank regularization, where
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{hard}(Z) 
  = 
  \lambda_\mathrm{spline} 
  \frac{1}{2} \left\|\Delta Z\right\|_2^2
  +
  \lambda_\mathrm{rank} \, \mathrm{rank}(Z).
\end{equation}
Here, the rank penalty is the actual rank of $Z$,
and note that $\mathrm{rank}(Z) = \| \sigma \|_0$.
In this case the problem is {\em not} convex.
However, we may still compute a proximal map for the non-differentiable rank term
and optimize with prox-gradients.

The third option allows more general smoothing but ignores all rank penalization:
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{indep}(Z) 
  = 
  \lambda_\mathrm{TV} 
  \left(
    \alpha_\mathrm{TV}\left\|\Delta Z\right\|_1 + \frac{(1-\alpha_\mathrm{TV})}{2}\left\|\Delta Z\right\|_2^2
  \right) .
\end{equation}
The TV terms control the smoothness of solutions, 
and the rank term allows us to recover low-rank solutions.
The function $R_\mathrm{indep}$ is convex; 
therefore we have a convex optimization problem.
Note also that the columns of $Z$ are uncoupled, 
so that we may solve this case as $\kappa_k$ different subproblems.

@@

>>>>

\title{Population genetics with evolving mutation processes: inference of mutation spectrum histories from modern genomic variation}

\author{
W.S. DeWitt$^{1,4}$, K. Decker Harris$^{2,3}$, and K. Harris$^{1,4}$\\
{\small
$^1$Department of Genome Sciences,
$^2$Paul G.\ Allen School of Computer Science \& Engineering,}\\
{\small and $^3$Department of Biology, University of Washington, Seattle, WA}\\
{\small $^4$Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA}
% \small{$^\ast$ Equal contribution}
}

@@

Models in evolutionary genetics typically assume that mutation rate is constant over time and between populations and closely related species.
However, recent work casts doubt on this assumption in human and ape populations, and reveals that mutation is a complex and dynamic process.
Whether arising from variation in replication fidelity, life history, or environmental exposures, mutation rate evolution can be accompanied by changes to the \emph{mutation spectrum}: the mutation rate in different local nucleotide contexts.
We extend theoretical tools based on Kingman's coalescent to accommodate a richly parameterized mutation process, varying in time and in spectrum.
We infer human mutation spectrum histories from patterns of modern genomic diversity, allowing us to reconstruct trajectories of divergence between populations, track a transient mutation spectrum perturbation through multiple populations, and characterize how mutation spectrum histories are structured by local nucleotide context.
Mutation rate and effective population size together determine the strength of genetic drift shaping genomic variation.
We introduce fast non-parametric joint inference of mutation spectrum history and demographic history.

@@

The first option we call ``soft'' rank regularization, with
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{soft}(Z)
  =
  \frac{  \lambda_\mathrm{spline} }{2} \left\|\Delta Z\right\|_2^2
  +
  \lambda_\mathrm{rank} \| Z \|_*
  +
   \frac{\lambda_\mathrm{ridge}}{2}  \| Z \|_\mathrm{F}^2
   .
\end{equation}
The spline term controls the smoothness of solutions,
and the rank term allows us to recover low-rank solutions.
The rank penalty is the nuclear norm $\| Z \|_*$,
which is equivalent to the $\ell^1$-norm on the singular values.
If $Z = U S V^\intercal$ is the SVD of $Z$, with $S = \mathrm{diag}(\sigma)$ then $\| Z \|_* = \| \sigma \|_1$.
The parameters $\lambda_\mathrm{spline}$ and $\lambda_\mathrm{rank}$ control the overall strength
of smoothing and rank regularization.
The function $R_\mathrm{soft}$ is convex; therefore we have a convex optimization problem
in this scenario.

The second option is ``hard'' rank regularization, where
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{hard}(Z)
  =
  \frac{\lambda_\mathrm{spline}}{2} \left\|\Delta Z\right\|_2^2
  +
  \lambda_\mathrm{rank} \, \mathrm{rank}(Z)
  +
  \frac{\lambda_\mathrm{ridge}}{2} \| Z \|_\mathrm{F}^2
  .
\end{equation}
Here, the rank penalty is the actual rank of $Z$,
and note that $\mathrm{rank}(Z) = \| \sigma \|_0$.
In this case the problem is {\em not} convex.
However, we may still compute a proximal map for the non-differentiable rank term
and optimize with prox-gradients.

The third option allows more general smoothing but ignores all rank penalization:
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{indep}(Z)
  =
  \lambda_\mathrm{TV}
  \left\|\Delta Z\right\|_1
  +
  \frac{\lambda_\mathrm{spline}}{2} \left\|\Delta Z\right\|_2^2
  +
  \frac{\lambda_\mathrm{ridge}}{2} \| Z \|_\mathrm{F}^2
  .
\end{equation}
The TV terms control the smoothness of solutions,
and the rank term allows us to recover low-rank solutions.
The function $R_\mathrm{indep}$ is convex.
Note also that the columns of $Z$ are uncoupled,
so that we may solve this case as $\kappa_k$ different subproblems.

@@

Erick Matsen, Peter Ralph, Andy Kern, Harris lab, Jeff Spence, Aleksandr Aravkin
